Multi-resolution Neural Field Estimation

### 2011-03-06 Multiresolution_framework/ltx/manuscript.tex
edits following mikes nd kens comments

\begin{abstract}
%\boldmath
The Integro-difference equation (IDE) is an increasingly popular model of spatio-temporal processes. Here we develop a multiresolution approximation (MRA) framework for the IDE neural field equations based on semi-orthogonal cardinal B-spline wavelets. State and parameter estimation is performed using the Expectation Maximisation (EM) algorithm. A synthetic example is given
to demonstrate the framework.
\end{abstract}
% Note that keywords are not normally used for peerreview papers.
% \begin{IEEEkeywords}
% Neural field model, multiresolution approximation (MRA), Expectation Maximization (EM) algorithm, wavelets.
% \end{IEEEkeywords}
% \IEEEpeerreviewmaketitle
\section{Introduction}
\IEEEPARstart{T}{he} human brain \mike{cortex?} has a multiresolution architecture, where spatial scales for information processing and transfer range from individual proteins in synapses\cut{, to neurons,} to networks of neurons. This multiresolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multiresolution data-driven framework for neural field modeling, the MultiResolution Approximate Integro-Difference Equation (MRAIDE), to address this challenge.

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. These models typically describe the macroscopic dynamics of the human brain, but are also descriptive of finer-scale neurodynamics. The ability to create patient-specific neural models will contribute to our knowledge base of poorly understood diseases such as epilepsy, and \cut{furthermore,} may also enable the development of new treatment strategies. This is particularly \cut{poignant with}\mike{relevant to} the advent of new devices that use therapeutic electrical stimulation. Presently, stimulation strategies for devices operate in an open-loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists an enormous potential to improve the performance of these devices using control theory,\mike{which in turn requires a suitable dynamic model.}\cut{ but first a suitable model is required.} \ken{I's rather not mentioning control when our model has no input.} Neural mass and neural field models are the ideal candidates for estimation and control algorithms, due to their strong links with the underlying physiology and their parsimony. It is expected that parameters of the neural fields models will be patient specific and will therefore, need to be inferred from data.

The first work describing data-driven mesoscopic neural modeling~\cite{Valdes1999} used a neural mass model~\cite{LopesDaSilva1976,Zetterberg1978} to fit EEG data. This approach was extended to coupled neural masses \cut{(using another model~}\cite{Jansen1995}) through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\cite{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explain the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently a framework was developed where a finite element model \parham{(FEM)} of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}. The current paper is an extension to this recent study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created where both macroscopic and microscopic behaviour of the system can be represented simultaneously.

@@

where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The term $e_t(\mathbf r)$ is an i.i.d. disturbance, such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$: a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$\cut{~\cite{Rasmussen2005}}. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})$. A linear form of the activation function was chosen to construct the estimator due to high computational demands when using a nonlinear (sigmoidal) activation function. \ken{Doing something just for computational efficiency only might be weak ... is there other work we can reference that makes the linearity assumption? ... or are there practical situations when this known to be a good approximations?} The observation equation describing the electrophysiological recordings is given by

@@

\section{Results}
To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically  using \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation}, enabling a comparison between true and estimated parameters. 200 realizations of 1 second of data were generated. The observation noise was set to $\boldsymbol\Sigma_{\epsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance was set to $\eta(r-r') = \phi_{3,-2}(r-r')$ (B-spline, $j=3$). The sampling period and the synaptic time constant were set to $T_s = 1$~ms and $\tau = 10$~ms, respectively. The firing rate slop, $\varsigma$ was set to $0.56  ~mV^{-1}$. The distance between adjacent sensors was $0.5$~mm resulting into $n_y = 161$ observations. The observation kernel was modeled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude. The spacing and bandwidth of the sensors allowed for the full spatial bandwidth of the field to be observed.  \ken{if we need to gain space I would remove the rest of this paragraph and put all parameters in a table} The spatial cutoff frequency of the observed field, $y_t(r_n)$, was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of \dean{6?}, the cutoff spatial frequency was \dean{$\nu_{cy} = ??$} cycles/mm \cite{Freestone2011}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states.

@@

Field reconstruction at different levels of approximations shown in \figurename{\ref{fig:FieldEstimates}}. The figure confirms that $j=3$ is adequate to capture the dynamics of the underlying field, where the RMSE converges to a steady value of 0.72. Fig.~\ref{fig:FieldEstimates} confirms good state estimation performance, where the estimated field is in good accordance with the true field.

The actual connectivity kernel and the decomposition is plotted in \figurename{\ref{fig:KernelEstimate}}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The small error in the estimate is attributed to the MRA of the system used to form the estimator.
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/Field.pdf}
		\caption{Actual (grey line) and estimated (black line) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$, $RMSE = 2.10$. (b) $j=1$, $n_x=33$, $RMSE = 1.42$. (c) $j=2$, $n_x=65$, $RMSE = 0.80$. (d) $j=3$, $n_x=131$, $RMSE = 0.72$. (e) $j=4$, $n_x=263$, $RMSE=0.72$.}
	\label{fig:FieldEstimates}
\end{figure} 
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/KernelUnknownSupportWithVarsigma.pdf}
		\caption{Connectivity kernel estimate. (a) The actual kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). (b) Kernel scaling functions at $j=1$. (c) Kernel wavelet functions at $j=1$.}
	\label{fig:KernelEstimate}
\end{figure}
% \caption{The actual connectivity kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). The 7 weighted kernel basis functions are shown by the black dashed lines.}
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution FEM \ken{(I might have missed it but did we define this acronym)} representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling. For example, it is thought that the dynamics and connectivity structure differs at different spatial scales~\cite{Qubbaj2009}.

@@

Future work should be directed towards extending the framework to account heterogeneity and perhaps time delays at lower spatial resolutions.\cut{ Another useful extensions would be the inclusion of a second-order synaptic response kernel to enrich the dynamics from the model.} Most importantly, future work should be directed towards applying and validating the framework on real data.

>>>>

\begin{abstract}
%\boldmath
The Integro-difference equation (IDE) is an increasingly popular model of spatio-temporal processes. Here we develop a multiresolution approximation (MRA) framework for the IDE neural field equations based on semi-orthogonal cardinal B-spline wavelets. State and parameter estimation is performed using the Expectation Maximisation (EM) algorithm. A synthetic example is given
to demonstrate the framework.
\end{abstract}
% Note that keywords are not normally used for peerreview papers.
% \begin{IEEEkeywords}
% Neural field model, multiresolution approximation (MRA), Expectation Maximization (EM) algorithm, wavelets.
% \end{IEEEkeywords}
% \IEEEpeerreviewmaketitle
\section{Introduction}
\IEEEPARstart{T}{he} human cerebral cortex has a multi-resolution architecture, where spatial scales for information processing and transfer range from individual proteins in synapses\cut{, to neurons,} to networks of neurons. This multi-resolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multi-resolution data-driven framework for neural field modeling, the Multi-Resolution Approximate Integro-Difference Equation (MRAIDE), to address this challenge.

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. These models typically describe the macroscopic dynamics of the human brain, but are also descriptive of finer-scale neurodynamics. The ability to create patient-specific neural models will contribute to our knowledge base of poorly understood diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Presently, stimulation strategies for devices operate in an open-loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists an enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are the ideal candidates for estimation algorithms, due to their strong links with the underlying physiology and their parsimony. It is expected that parameters of the neural fields models will be patient specific and will therefore, need to be inferred from data.

The first work describing data-driven mesoscopic neural modeling used a neural mass model to fit EEG data~\cite{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\cite{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explain the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently a framework was developed where a finite element model (FEM) of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}. The current paper is an extension to this recent study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created where both macroscopic and microscopic behaviour of the system can be represented simultaneously.

@@

where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The term $e_t(\mathbf r)$ is an i.i.d. disturbance, such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$: a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})$. A linear form of the activation function was chosen to construct the estimator due to high computational demands when using a nonlinear (sigmoidal) activation function. \ken{Doing something just for computational efficiency only might be weak ... is there other work we can reference that makes the linearity assumption? ... or are there practical situations when this known to be a good approximations?} The observation equation describing the electrophysiological recordings is given by

@@

\section{Results}
To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically  using \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation}, enabling a comparison between true and estimated parameters. 200 realizations of 1 second of data were generated. The observation noise was set to $\boldsymbol\Sigma_{\epsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance was set to $\eta(r-r') = \phi_{3,-2}(r-r')$ (B-spline, $j=3$). The sampling period and the synaptic time constant were set to $T_s = 1$~ms and $\tau = 10$~ms, respectively. The firing rate slop, $\varsigma$ was set to $0.56~mV^{-1}$. The distance between adjacent sensors was $0.5$~mm, resulting into $n_y = 161$ observations. The observation kernel was modeled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude. The spacing and bandwidth of the sensors allowed for the full spatial bandwidth of the field to be observed.  \ken{if we need to gain space I would remove the rest of this paragraph and put all parameters in a table} The spatial cutoff frequency of the observed field, $y_t(r_n)$, was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of \dean{6?}, the cutoff spatial frequency was \dean{$\nu_{cy} = ??$} cycles/mm \cite{Freestone2011}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states.

@@

The field reconstruction, at different levels of approximations, is shown in \figurename{\ref{fig:FieldEstimates}}. The figure confirms that $j=3$ is adequate to capture the dynamics of the underlying field, where the RMSE converges to a steady value of 0.72. Fig.~\ref{fig:FieldEstimates} confirms good state estimation performance, where the estimated field is in good accordance with the true field.

The actual connectivity kernel and the decomposition is plotted in \figurename{\ref{fig:KernelEstimate}}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The small error in the estimate is attributed to the MRA of the system used to form the estimator.
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/Field.pdf}
		\caption{Actual (grey line) and estimated (black line) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$, $RMSE = 2.10$. (b) $j=1$, $n_x=33$, $RMSE = 1.42$. (c) $j=2$, $n_x=65$, $RMSE = 0.80$. (d) $j=3$, $n_x=131$, $RMSE = 0.72$. (e) $j=4$, $n_x=263$, $RMSE=0.72$.}
	\label{fig:FieldEstimates}
\end{figure} 
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/KernelUnknownSupportWithVarsigma.pdf}
		\caption{Connectivity kernel estimate. (a) The actual kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). (b) Kernel scaling functions at $j=1$. (c) Kernel wavelet functions at $j=1$.}
	\label{fig:KernelEstimate}
\end{figure}
% \caption{The actual connectivity kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). The 7 weighted kernel basis functions are shown by the black dashed lines.}
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution FEM representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling. For example, it is thought that the dynamics and connectivity structure differs at different spatial scales~\cite{Qubbaj2009}.

@@

Future work should be directed towards extending the framework to account for a second-order synaptic response kernel, heterogeneity in the connectivity and perhaps time delays at lower spatial resolutions. Most importantly, future work should be directed towards applying and validating the framework on real data.
